On the power of conditional independence testing \\ under model-X

### 2022-05-04 arxiv-v4/crtpower.tex
Fixed typos found by Aaditya

\usepackage[uniquename=false, uniquelist = false, url = false, doi = false, isbn = false, sorting = none, natbib = true, backend = bibtex]{biblatex}

@@

In this section, we seek the most powerful MX CI test against a point alternative. To accomplish this, we observe that any (marginally) level $\alpha$ test must also have level $\alpha$ conditionally on $\sry, \srz$ (Section~\ref{sec:reduction-by-conditioning}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to Appendix~\ref{sec:proofs-sec2}.

@@

\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\nonl  \textbf{Input:} $\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, distribution  $f^*_{\prx|\prz}$, distillation function $d$, test statistic $T_n^d$, number of resamples $B$\\
			Distill information in $\srz$ about $\srx$ and $\sry$ into $D \equiv d(\sry, \srz)$\;
			\For{$b = 1, 2, \dots, B$}{
				Resample $\srxk_i^{(b)} \overset{\text{ind}}\sim f^*_{\prx|\prz = \srz_i},\ i = 1, \dots, n$\;
			}
			Compute $\widehat p \equiv \frac{1}{B+1}\sum_{b = 1}^B \mathbbm 1(T_n^d(\srxk^{(b)}, \sry, D) \geq T_n^d(\srx, \sry, D)$.\\
			\nonl \textbf{Output:} dCRT $p$-value $\widehat p$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit, and drawing $B$ resamples.
			\caption{\bf The distilled conditional randomization test}
			\label{alg:dCRT}
		\end{algorithm}
	\end{minipage}
\end{center}

@@

Under Setting~\ref{setting:orthogonal-design}, we can directly use Bayati and Montanari's theory \cite{Bayati2011} to obtain 
\begin{equation}
\lim_{n \rightarrow \infty}\mathcal E_n^2 = \tau_*^2 - \sigma^2 \quad \text{a.s. in } (\srx_{\text{train}}, \sry_{\text{train}}, \srz_{\text{train}}),
\label{eq:bm-result}
\end{equation}
where $(\alpha_*,\tau_*)$ is the unique solution of the system below:
\begin{equation}
	\begin{split}
		\lambda &= \alpha \tau (1-(\pi\delta)^{-1}\mathbb E[\eta'(\sqrt \pi \Gamma + \tau W; \alpha \tau)]) \\
		\tau^2 &= \sigma^2 + (\pi\delta)^{-1}\mathbb E[(\eta(\sqrt \pi\Gamma + \tau W; \alpha \tau) - \sqrt \pi\Gamma)^2].
	\end{split}
	\label{eq:amp-system}
\end{equation}
Here, $W \sim N(0,1)$ is independent of $\Gamma$ and $\eta(x; \theta) = (|x|-\theta)_+\textnormal{sign}(x)$ is the soft threshold function. This leads to the following corollary of Theorem~\ref{thm:power}, proved in Appendix~\ref{sec:proofs-sec45}:
\begin{corollary} \label{cor:lasso}
Under Setting~\ref{setting:orthogonal-design}, the asymptotic power of the dCRT converges to that of a standard normal location test with alternative mean $\tau_*^{-1} h\sqrt{1-\pi}$:
\begin{equation}
	\begin{split}
	\lim_{n \rightarrow \infty}\mathbb E_{\mathcal L_n}[\varphi^{\mathcal L_n}_{n}(\srx, \sry, \srz)] =  \mathbb P[|N(\tau_*^{-1} h\sqrt{1-\pi},1)| > z_{1-\alpha/2}].
	\label{eq:lasso-power}
	\end{split}
\end{equation}
\end{corollary}
Corollary~\ref{cor:lasso} gives the power of these lasso-based methods in a very simple form, with the prediction error of the lasso entering through the effective noise level $\tau_*$. The impact of the splitting proportion $\pi$ on power can be seen in the multiplication of the signal strength $h$ by $\sqrt{1-\pi}$. The splitting proportion implicitly impacts the effective noise level $\tau_*$ as well; smaller $\pi$ lead to greater effective noise levels. Note that the expectations in Corollary~\ref{cor:lasso} are over both training and test sets, while the expectations in Theorem~\ref{thm:power} are over the test set only.

@@

\begin{proof}[Proof of Theorem~\ref{prop:crt-optimality}]
	Fix realizations $\sfy, \sfz$. We first claim that $\phi^{\CRT}_{T^{\textnormal{opt}}}$ is the most powerful test in the conditional problem, i.e.
	\begin{equation}
		\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sfy,\sfz)|\sry = \sfy, \srz = \sfz] \leq \mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx, \sfy, \sfz)|\sry = \sfy, \srz = \sfz]
		\label{conditional-optimality}
	\end{equation}
	for any test $\phi(\srx, \sfy, \sfz)$ satisfying
	\begin{equation}
		\sup_{\mathcal L \in \mathscr L_0^{\text{MX}}(f^*)}\mathbb E_{\mathcal L}[\phi(\srx,\sfy,\sfz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
		\label{eq:conditional-level}
	\end{equation}
	In the conditional problem, the alternative $\bar{\mathcal L}$ induces the following distribution for $\srx$:
	\begin{equation}
		\begin{split}
			\bar{\mathcal L}(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i|\sfz_i)\tfrac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)}
			\label{conditional-alternative}
		\end{split}
	\end{equation}
	where
	\begin{equation*}
	\bar f(\sfy_i|\sfz_i) \equiv \int \bar f(\sfy_i|\sfx_i, \sfz_i)f^*(\sfx_i|\sfz_i)d\sfx_i.
	\end{equation*}	
	The conditional problem is therefore a test of 
	\begin{equation*}
		\begin{split}
			&H_0: \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i) \quad \text{versus} \\
			&H_1: \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i|\sfz_i)\tfrac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)}.
		\end{split}
	\end{equation*}
	This is a simple testing problem, with point null and point alternative. By the Neyman-Pearson lemma, the most powerful test is the one that rejects for large values of the likelihood ratio
	\begin{equation}
		\prod_{i = 1}^n \frac{P_1(\sfx_i|\sfy_i, \sfz_i)}{P_0(\sfx_i|\sfy_i, \sfz_i)} = \prod_{i = 1}^n \frac{f^*(\sfx_i|\sfz_i)\frac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)}}{f^*(\sfx_i|\sfz_i)} = \prod_{i = 1}^n \frac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)} \propto T^{\text{opt}}(\sfx, \sfy, \sfz),
	\label{eq:likelihood-ratio-derivation}
	\end{equation}
	verifying the conditional optimality claim~\eqref{conditional-optimality}. To obtain the unconditional optimality claim~\eqref{unconditional}, note that by Proposition~\ref{prop:marginal-implies-conditional} any unconditionally level $\alpha$ test $\phi$ must also have level $\alpha$ in the conditional problem~\eqref{eq:conditional-level}. For any such test, we may therefore conclude
	\begin{equation*}
		\begin{split}
			\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sry,\srz)] &= \mathbb E_{\bar{\mathcal L}}[\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sry,\srz)|\sry,\srz]] \\
			&\leq \mathbb E_{\bar{\mathcal L}}\left[\mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx,\sry,\srz)|\sry,\srz]\right] = \mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx, \sry, \srz)],
		\end{split}
	\end{equation*}
as desired.
\end{proof}

>>>>

\usepackage[uniquename=false, uniquelist = false, url = false, doi = false, isbn = false, sorting = none, natbib = true, backend = bibtex, maxbibnames=10]{biblatex}

@@

In this section, we seek the most powerful MX CI test against a point alternative. To accomplish this, we observe that any (marginally) level $\alpha$ test must also have level $\alpha$ conditionally on $\sry, \srz$ (Section~\ref{sec:reduction-by-conditioning}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman-Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to Appendix~\ref{sec:proofs-sec2}.

@@

\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\nonl  \textbf{Input:} $\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, distribution  $f^*_{\prx|\prz}$, distillation function $d$, test statistic $T_n^d$, number of resamples $B$\\
			Distill information in $\srz$ about $\srx$ and $\sry$ into $D \equiv d(\sry, \srz)$\;
			\For{$b = 1, 2, \dots, B$}{
				Resample $\srxk_i^{(b)} \overset{\text{ind}}\sim f^*_{\prx|\prz = \srz_i},\ i = 1, \dots, n$\;
			}
			Compute $\widehat p \equiv \frac{1}{B+1}\sum_{b = 1}^B \mathbbm 1(T_n^d(\srxk^{(b)}, \sry, D) \geq T_n^d(\srx, \sry, D))$.\\
			\nonl \textbf{Output:} dCRT $p$-value $\widehat p$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit, and drawing $B$ resamples.
			\caption{\bf The distilled conditional randomization test}
			\label{alg:dCRT}
		\end{algorithm}
	\end{minipage}
\end{center}

@@

Under Setting~\ref{setting:orthogonal-design}, we can directly use Bayati and Montanari's theory \cite{Bayati2011} to obtain 
\begin{equation}
\lim_{n \rightarrow \infty}\mathcal E_n^2 = \tau_*^2 - \sigma^2 \quad \text{a.s. in } (\srx_{\text{train}}, \sry_{\text{train}}, \srz_{\text{train}}),
\label{eq:bm-result}
\end{equation}
where $(\alpha_*,\tau_*)$ is the unique solution of the system below:
\begin{equation}
	\begin{split}
		\lambda &= \alpha \tau (1-(\pi\delta)^{-1}\mathbb E[\eta'(\sqrt \pi \Gamma + \tau W; \alpha \tau)]), \\
		\tau^2 &= \sigma^2 + (\pi\delta)^{-1}\mathbb E[(\eta(\sqrt \pi\Gamma + \tau W; \alpha \tau) - \sqrt \pi\Gamma)^2].
	\end{split}
	\label{eq:amp-system}
\end{equation}
Here, $W \sim N(0,1)$ is independent of $\Gamma$ and $\eta(x; \theta) = (|x|-\theta)_+\textnormal{sign}(x)$ is the soft threshold function. This leads to the following corollary of Theorem~\ref{thm:power}, proved in Appendix~\ref{sec:proofs-sec45}:
\begin{corollary} \label{cor:lasso}
Under Setting~\ref{setting:orthogonal-design}, the asymptotic power of the dCRT converges to that of a standard normal location test with alternative mean $\tau_*^{-1} h\sqrt{1-\pi}$:
\begin{equation}
	\begin{split}
	\lim_{n \rightarrow \infty}\mathbb E_{\mathcal L_n}[\varphi^{\mathcal L_n}_{n}(\srx, \sry, \srz)] =  \mathbb P[|N(\tau_*^{-1} h\sqrt{1-\pi},1)| > z_{1-\alpha/2}].
	\label{eq:lasso-power}
	\end{split}
\end{equation}
\end{corollary}
Corollary~\ref{cor:lasso} gives the power of these lasso-based methods in a very simple form, with the prediction error of the lasso entering through the effective noise level $\tau_*$. The impact of the splitting proportion $\pi$ on power can be seen in the multiplication of the signal strength $h$ by $\sqrt{1-\pi}$. The splitting proportion implicitly impacts the effective noise level $\tau_*$ as well; smaller $\pi$ lead to greater effective noise levels. Note that the expectations in Corollary~\ref{cor:lasso} are over both training and test sets, while the expectations in Theorem~\ref{thm:power} are over the test set only.

@@

\begin{proof}[Proof of Theorem~\ref{prop:crt-optimality}]
	Fix realizations $\sfy, \sfz$. We first claim that $\phi^{\CRT}_{T^{\textnormal{opt}}}$ is the most powerful test in the conditional problem, i.e.
	\begin{equation}
		\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sfy,\sfz)|\sry = \sfy, \srz = \sfz] \leq \mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx, \sfy, \sfz)|\sry = \sfy, \srz = \sfz]
		\label{conditional-optimality}
	\end{equation}
	for any test $\phi(\srx, \sfy, \sfz)$ satisfying
	\begin{equation}
		\sup_{\mathcal L \in \mathscr L_0^{\text{MX}}(f^*)}\mathbb E_{\mathcal L}[\phi(\srx,\sfy,\sfz)|\sry = \sfy, \srz = \sfz] \leq \alpha.
		\label{eq:conditional-level}
	\end{equation}
	In the conditional problem, the alternative $\bar{\mathcal L}$ induces the following distribution for $\srx$:
	\begin{equation}
		\begin{split}
			\bar{\mathcal L}(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i|\sfz_i)\tfrac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)},
			\label{conditional-alternative}
		\end{split}
	\end{equation}
	where
	\begin{equation*}
	\bar f(\sfy_i|\sfz_i) \equiv \int \bar f(\sfy_i|\sfx_i, \sfz_i)f^*(\sfx_i|\sfz_i)d\sfx_i.
	\end{equation*}	
	The conditional problem is therefore a test of 
	\begin{equation*}
		\begin{split}
			&H_0: \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i) \quad \text{versus} \\
			&H_1: \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i|\sfz_i)\tfrac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)}.
		\end{split}
	\end{equation*}
	This is a simple testing problem, with point null and point alternative. By the Neyman-Pearson lemma, the most powerful test is the one that rejects for large values of the likelihood ratio
	\begin{equation}
		\prod_{i = 1}^n \frac{P_1(\sfx_i|\sfy_i, \sfz_i)}{P_0(\sfx_i|\sfy_i, \sfz_i)} = \prod_{i = 1}^n \frac{f^*(\sfx_i|\sfz_i)\frac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)}}{f^*(\sfx_i|\sfz_i)} = \prod_{i = 1}^n \frac{\bar f(\sfy_i|\sfx_i, \sfz_i)}{\bar f(\sfy_i|\sfz_i)} \propto T^{\text{opt}}(\sfx, \sfy, \sfz),
	\label{eq:likelihood-ratio-derivation}
	\end{equation}
	verifying the conditional optimality claim~\eqref{conditional-optimality}. To obtain the unconditional optimality claim~\eqref{unconditional}, note that by Proposition~\ref{prop:marginal-implies-conditional} any unconditionally level $\alpha$ test $\phi$ must also have level $\alpha$ in the conditional problem~\eqref{eq:conditional-level}. For any such test, we may therefore conclude
	\begin{equation*}
		\begin{split}
			\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sry,\srz)] &= \mathbb E_{\bar{\mathcal L}}[\mathbb E_{\bar{\mathcal L}}[\phi(\srx,\sry,\srz)|\sry,\srz]] \\
			&\leq \mathbb E_{\bar{\mathcal L}}\left[\mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx,\sry,\srz)|\sry,\srz]\right] = \mathbb E_{\bar{\mathcal L}}[\phi^{\CRT}_{T^\textnormal{opt}}(\srx, \sry, \srz)],
		\end{split}
	\end{equation*}
as desired.
\end{proof}
